Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology

### 2023-06-17 manuscript/manuscript.qmd
quarto formatting

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
preprint: true
author:
  - name: "Anton Olsson-Collentine"
    affiliation: 1
    corresponding: true
    email: anton@olssoncollentine.com

  - name          : "Marjan Bakker"
    affiliation   : "1"

  - name          : "Jelte Wicherts"
    affiliation   : "1"
affiliation:
  - code: 1
    address: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."

abstract: >
  One of the main purposes of meta-analysis is to estimate heterogeneity of effect sizes. The quality of meta-analytic estimates depends in large part on the quality of evidence in primary studies, yet there exist broad concerns that proper measurement tends to be neglected in much of psychological primary research. Such inattention to measurement can also bias heterogeneity estimates. We performed a Monte-Carlo simulation study to examine the expected bias in heterogeneity estimates due to unreliability in primary studies. Although unreliability in primary studies can both inflate and suppress heterogeneity, our results indicate that under most circumstances unreliability in primary studies leads to a severe underestimate of heterogeneity in meta-analyses. For a typical effect size heterogeneity can be expected to be underestimated by 15-60%, or more if average reliability in primary studies is below 0.6. We observed a positive bias in heterogeneity estimates due to unreliability only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. The severity of the underestimation of heterogeneity could lead to meta-analyses in psychology missing relevant moderators and overgeneralizing outcomes. Unreliability appears to bring truly different effects closer together, thereby obscuring true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is urgently needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.

header-includes:
   - \usepackage{setspace}
   - \usepackage{lineno}
   - \linenumbers

csl: apa.csl

output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    keep_tex: true
  rticles::peerj_article: default

---

@@

Table 1.

*Variance in reliability inflates heterogeneity in combination with increased average effect size and leads to overestimated heterogeneity in a fixed effects model*

@@

@@

Table 2.

*Imperfect reliability suppresses true heterogeneity*

@@

@@

```{r Figure 1, out.width= "100%"}
knitr::include_graphics("../figures/manuscript/flowchart-simulation-design.png")
```

_Figure 1. Flowchart summarizing the simulation design. The flowchart shows the procedures followed to meta-anlyze results using the Hedge’s and Vevea model (see ‘Meta-Analytic Model’ header in methods section). The Hunter & Schmidt ‘bare-bones’ meta-analytic model followed the Pearson correlations procedure, but used sample size weights and replaced_ $r_i$ _in the sampling variance by $\bar{r}$ for meta-anlysis._

>>>>

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
date: today
date-format: long
author:
  - name: Anton Olsson-Collentine
    url: https://example.com
    orcid: 0000-0000-0000-0000
    affiliations:
      - id: tiu
        name: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."
    email: anton@olssoncollentine.com
  - name: Marjan Bakker
    affiliations:
      - ref: tiu
  - name: Jelte Wicherts
    affiliations:
      - ref: tiu
format:
  bj-pdf:
    classoption:
      - twocolumn
colorlinks: true
link-citations: true
keep-tex: false
abstract: |
  One of the main purposes of meta-analysis is to estimate heterogeneity of effect sizes. The quality of meta-analytic estimates depends in large part on the quality of evidence in primary studies, yet there exist broad concerns that proper measurement tends to be neglected in much of psychological primary research. Such inattention to measurement can also bias heterogeneity estimates. We performed a Monte-Carlo simulation study to examine the expected bias in heterogeneity estimates due to unreliability in primary studies. Although unreliability in primary studies can both inflate and suppress heterogeneity, our results indicate that under most circumstances unreliability in primary studies leads to a severe underestimate of heterogeneity in meta-analyses. For a typical effect size heterogeneity can be expected to be underestimated by 15-60%, or more if average reliability in primary studies is below 0.6. We observed a positive bias in heterogeneity estimates due to unreliability only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. The severity of the underestimation of heterogeneity could lead to meta-analyses in psychology missing relevant moderators and overgeneralizing outcomes. Unreliability appears to bring truly different effects closer together, thereby obscuring true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is urgently needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.
license: CC BY
keywords:
  - Research
  - Cool Stuff
citation:
  type: article
  citation-key: citationKeyNumber
  status: draft
cite-method: citeproc
csl: apa.csl
bibliography: references.bib
---

@@

@@

#| tbl-pos: t

@@

@@

```{r Figure 1, out.width= "85%"}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Figure 1. Flowchart summarizing the simulation design. The flowchart shows the procedures followed to meta-anlyze results using the Hedge’s and Vevea model (see ‘Meta-Analytic Model’ header in methods section). The Hunter & Schmidt ‘bare-bones’ meta-analytic model followed the Pearson correlations procedure, but used sample size weights and replaced $r_i$ in the sampling variance by r for meta-anlysis."

knitr::include_graphics("../figures/manuscript/flowchart-simulation-design.png")
```

@@
